Variation in gaze understanding across the life span: A process-level perspective

### 2023-09-01 paper/manuscript.Rmd
major edits in intro & discussion

abstract: |
  Abstract max 250 words

keywords          : "social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan"
wordcount         : "xxx max 8000"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes   :
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
---

@@

**Acknowledgements:** We thank Jana Jurkat for her help with data collection and participant recruitment.
We would also like to thank Steven Kalinke for his technical programming support.

**Funding:** This study was funded by the Max Planck Society for the Advancement of Science, a noncommercial, publicly financed scientific organization (no grant number).
Manuel Bohn was supported by a Jacobs Foundation Research Fellowship.
We thank all the children, caregivers, and adults who participated in the study.

**Ethical statement:** The studies were approved by an internal ethics committee at the Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany.Informed consent was obtained from all individual participants or their legal guardians.

@@

# Research highlights

-   up to four bulleted points outlining the key contributions to research the paper makes.
-   The Research Highlights should be placed before the abstract.
-   Each research highlight should not be longer than 25 words.

# Introduction

-   why do we care about developmental trajectory? ref to stat learning paper
-   variation

## Why do we need gaze understanding?

How do humans learn about their environment and navigate through their social surroundings?
One possibility to extract information from the environment is through following others' focus of attention.
Building a common ground is considered especially important in communicative interactions and shared activities [@tomasello2007reliance].

## How does gaze following emerge?

Existing studies operationalize gaze following as the ability to follow another agent's line of sight.
As one of the most fundamental social-cognitive abilities, it has been extensively studied in infancy and early childhood.
Infants as young as six months can attune their gaze to that of another agent [@dentremont1997demonstration].
At the end of their first year of life, infants can follow gaze to locations outside their current visual field and move themselves to gain proper perceptual access [@moll200412].

While the emergence of gaze following has been well established, less is known about the developmental trajectory throughout childhood and adolescence.
One possibility is that our social-cognitive ability in question is fully developed once emerged in infancy.
However, many cognitive abilities develop with age [e.g., working memory, @gathercole2004structure].
Similarly, visual processing appears to improve with age.
Therefore, children could potentially improve in gaze following, fine-tuning the performance of the already existing skill.

## The scope of infants' gaze following ability

Though these studies suggest that young infants can align their visual attention to another's line of sight, it does not necessarily include understanding the intentions of the other agent.
Infants could simply attune their orientation or be attracted by others' gaze without processing what exactly the other is seeing (cf. Butterworth & Jarrett's ecological and geometric mechanism, @butterworth1991minds]. Therefore, it is crucial to study children's intentional understanding of gaze.

@moore1997role showed that 9-month-olds followed an agent's gaze more, when it was accompanied by a dynamic head turn in comparison to a static head turn.

In a hiding game with two search locations, @povinelli1997exploitation found that three-year-olds used gaze as a cue to locate the reward, while two-year-olds performed at chance level.

In a similar object choice paradigm with two containers, @behne2005oneyearolds investigated whether infants understand the communicative intent behind pointing and gaze cues.
In contrast to @povinelli1997exploitation, they found that already 14-month-olds used the agent's cues to select an object.
In conditions with absent-minded 'cues', infants performed around chance.
This could be interpreted as infants recognizing the nature of this joint activity: namely, that the adult's behavior was beneficial and relevant for their object choice.

### Head vs eye direction

It is important to note that in many existing gaze conditions, the experimenter shifted their eyes and head in synchrony (e.g., [@povinelli1997exploitation, @behne2005oneyearolds]).
Instead of pointing towards gaze understanding, a critic could claim that the results can be explained by face direction alone.

A handful of studies approached this potential confound by separately manipulating head and eye movement.
@brooks2002importance implemented a comparison between eye and head orientation and found that 14-month-olds were sensitive to open versus closed eyes.

Investigating the 'cooperative eye hypothesis', @tomasello2007reliance implemented six conditions, in which an experimenter oriented towards the ceiling with their eyes only, head only (eyes closed), both head and eyes, or neither.
They found that human infants relied more on the eye movement, while chimpanzees paid more attention to the head movement.

Importantly, the subjects were not presented with an object choice but their attention orientation was measured.

-   [@raviv2018developmental]

-   [@astor2022gaze]

-   [@colombo2001development]

-   [@scaife1975capacity]

-   [@itakura1998use]

-   [@carpenter1998social] "Several other studies have attempted to determine more precisely the cue that infants are using when they follow the gaze direction of others, that is, whether they use adults' head or eye orientation. In tasks comparing infants' responses when the experimenters turned their head and eyes together to targets with their responses when the experimenters directed their eyes to the targets but their head remained facing forward, Corkum and Moore (1995), Lempers (1979), and Lempers, Flavell, and Flavell (1977) all found that only infants age 12 months and older responded correctly when eyes and head were oriented in the same direction and that infants at all ages (i.e., through 19 months) performed poorly when eye and head direction diverged" (p.10-11) object choice.

-   [@silverstein2021infants] for vertical plane

-   [@zhang2019role]

-   [@frischen2007gaze]

-   [@lee1998children]

-   [@coelho2006searching]

@@

### Developmental trajectory, measuring & modeling individual differences

In this study, we were interested in the developmental trajectory of gaze understanding.
While we expect the younger children to be able to follow gaze, we aimed at assessing the differentiation of their social-cognitive ability.
Our goal was *not* to establish the youngest age at which children understand gaze cues.
Rather, we wanted to examine how that ability changes with age.

In our study, we focused on the communicative intents of gaze: we asked children to locate a target by following an agent's gaze.
While language demands were kept low, the participants had to actively respond and, therefore, make use of the presented gaze cue.

A unique contribution of this study is the richness of the data set.
Methodological challenges arise when trying to compare data across ages from qualitatively and quantitatively different study tasks.
We could circumvent these issues by applying the exact same task for the entire life span.

@@

<https://osf.io/6yjz3>

@@

We collected data from a remote child, teenager and adult sample, reaching from 3 to 80 years of age (see Supplements for further details).

The child and teenager sample consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group < 18])` participants.
We recruited participants via an internal database consisting of families living in Leipzig, Germany, who volunteered to participate in child development studies and indicated an interest in online studies.
Participants came from ethnically homogeneous, mixed socioeconomic backgrounds with mid to high parental education levels.
They lived in an industrialized, urban Central-European context in a mid-size German city (∼600,000 inhabitants; median individual monthly net income \~ 1,600€ as of 2021).
Most were raised monolingually in a nuclear two-generational family setting.
Information on demographics and socioeconomic status was not formally recorded on a participant level.

Adults were recruited via *Prolific* [@palan2018prolific].
*Prolific* is an online participant recruitment service from the University of Oxford with a predominantly European and US-American subject pool.
Participants consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group >= 20])` English-speaking adults that reported to have normal or corrected-to-normal vision.
For completing the study, subjects were paid above the fixed minimum wage (on average £10.00 per hour).
For further information on age and gender of participants, see Table 1.

@@

We used the continuous version of the TANGO [@prein2023tango].
The task was presented as an interactive web application (see Figure \@ref(fig:fig1); live demo [https://ccp-odc.eva.mpg.de/tango-demo/](https://ccp-odc.eva.mpg.de/tango-demo/.); source code <https://github.com/ccp-eva/tango-demo>).
The TANGO showed satisfactory internal consistency and retest reliability [with reliability estimates *Pearson's r* ranging from .7 to .8 for the continuous task version; @prein2023tango].

@@

To estimate the effect of age on gaze understanding, we fit GLMMs that make different assumptions about the developmental trajectory, modeling the relationship as linear, quadratic, or cubic.
In addition, we applied a Gaussian Process model [@burkner2017brms] and examined nonlinearity.
The models predicted imprecision by age (continuous), aggregated across trials and modeled as a `lognormal` distribution.[^1]
The unit of imprecision was counted in target width, i.e., a participant with imprecision of 1 clicked on average one target width to the left or right of the true target center.
We inspected the posterior distributions (mean and 95% Credible Interval (CrI)) for the age estimates and compared models using leave-one-out cross-validation (LOO), the widely applicable information criterion (WAIC), and model weights [@vehtari2017practicala].

[^1]: Originally, our models were fitted on a trial-by-trial basis and implemented the following structure: `performance ~ age + symmetricPosition + trialNr + (1 + symmetricPosition + trialNr | subjID)`.
    However, the Gaussian Process model was computationally heavy.
    To circumvent issues of convergence and for practicability, we simplified the model structure, aggregated data on a subject level, and included only age as an effect.
    In the Supplements, you can find a comparison between the original and the here reported model structures.
    Essentially, the model predictions did not differ notably.

@@

(ref:figlab1) **Developmental trajectory of gaze understanding across the lifespan**. Performance is measured as imprecision, i.e., the absolute distance between the target's center and the participant's click (averaged across trials). The unit of imprecision is counted in the width of the target, i.e., a participant with imprecision of 1 clicked on average one target width to the left or right of the true target center. Grey dots show the mean performance for each subject averaged across trials. Blue lines show 100 draws from the expectation of the posterior predictive distribution (i.e, conditional expectation) of the Gaussian Process model, with its mean predicted developmental trajectory as a solid black line. Vertical, black, dashed lines show the location of the most prominent change points.

```{r fig1, include = T, out.width = "100%", fig.align = "center", fig.cap = "(ref:figlab2)"}
knitr::include_graphics(c("../figures/lifespan_all.png"))
```

@@

High levels of variation pointed to individual differences in all age groups (overall imprecision mean = `r lifespan_variation$imprecision_mean`, sd = `r lifespan_variation$imprecision_sd`, range = [`r lifespan_variation$imprecision_min` - `r lifespan_variation$imprecision_max`]).
For example, there were some children that were more accurate than the average adult.

In our model comparison, we found clear evidence for a non-linear development in gaze understanding across the lifespan.
Compared against a linear, quadratic, and cubic model, the Gaussian Process model showed the best model fit, demonstrated by the greatest model weight and smallest WAIC values.
For the imprecision in gaze understanding, the standard deviation (*b* = `r lifespan_gp$estimate[lifespan_gp$param == "sdgp(gpage_centered)"]`, 95% CrI [`r lifespan_gp$l_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`; `r lifespan_gp$u_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`]) and visual inspection indicated nonlinearity (see Figure \@ref(fig:fig2)).

Going one step further, we investigated the most prominent change points in the data.
The Bayesian change point analysis revealed `r lifespan_beast$trend$ncp %>% round(1)` (with `r lifespan_beast$trend$ncpPr[lifespan_beast$trend$ncp %>% round(1) + 1] * 100 %>% round(2)`% probability) major shifts in gaze understanding during the lifespan.
The change points occurred at the following ages: `r lifespan_cps$cp[1]` years (95% CrI [`r lifespan_cps$lowerCrI[1]`; `r lifespan_cps$upperCrI[1]`]); followed by `r lifespan_cps$cp[2]` years (95% CrI [`r lifespan_cps$lowerCrI[2]`; `r lifespan_cps$upperCrI[2]`]); followed by `r lifespan_cps$cp[3]` years (95% CrI [`r lifespan_cps$lowerCrI[3]`; `r lifespan_cps$upperCrI[3]`]); and finally at `r lifespan_cps$cp[4]` years (95% CrI [`r lifespan_cps$lowerCrI[4]`; `r lifespan_cps$upperCrI[4]`]).
In other words: we found a very rapid initial improvement in early childhood (three change points in rapid succession), followed by a long period of minor, very slow change with slightly increasing levels of imprecision toward the eldest in our sample.

@@

We investigated the shape of change in gaze understanding across the lifespan.
By applying the exact same task for the entire life span, we could directly compare gaze understanding in all ages.
Therefore, we circumvented methodological challenges that often arise when trying to compare data across ages from qualitatively and quantitatively different study tasks.

@@

Previous studies found that children start to follow gaze in the second half of their first year of life [@moll200412].
In our sample, three-year-olds were still rather imprecise in their gaze understanding ability.
How can we explain this divergence?
First of all, we used subtle eye movements as cues.
Many existing studies let the agents move eye and head in parallel [@behne2005oneyearolds; @povinelli1997exploitation], therefore establishing a confound with greater (head) movement.
Relying exclusively on eye movements might be more difficult for children than presenting them with a combined eye and head orientation [@carpenter1998social].
Furthermore, our study required participants to (1) precisely follow an agent's gaze, (2) interpret this as a cue, and then (3) make use of this cue to guide their own behavior (i.e., touching the screen at the cued location).
We measured an active location choice instead of an attention orientation.
It is conceivable that three-year-olds followed the agent's gaze but were still learning to translate this understanding into precise, active behavior.

Regarding the sample of elderly adults, we expect a sampling bias to be present [@bethlehem2010selection; @gosling2004should; @remillard2014systematic].
First, certainly not all older people have working a high-speed Internet connection or are knowledgeable and trained in its use.
Second, it takes a certain amount of independence and motivation to participate in *Prolific* studies.
The elderly who know how to participate in *Prolific* studies certainly might show greater cognitive fitness and flexibility compared to their offline counterparts.
Therefore, a representative sample might show a greater age decline in gaze understanding compared to our reported sample.
In addition, it is noteworthy that older people might be more likely to suffer from visual impairments.
Even though we filtered participants to only include normal- to correct-to-normal vision, we cannot guarantee that our participants showed no symptoms of reduced vision.

@@

Our lifespan study showed that gaze understanding develops throughout childhood, and variation between individuals appears in all age groups.
The TANGO has previously been shown to be reliably capture inter-individual differences in gaze understanding [@prein2023tango].
The variation between participants is thus likely genuine and not due to random noise.
Now, we aim to understand what explains the developmental change and the variation across participants on a process level.
We present a theory of gaze understanding that explains how children process the available information (i.e., the agent's eyes) to make inferences about the agent's gaze and its attentional focus, which leads them to identify the target position.
We formalize this inference process in a computational cognitive model.

@@

We seek to explain how participants solved the TANGO.
Our gaze model assumes that participants use available gaze information to infer the target location.

@@

In addition, we included `r gazemodel_sample$n_total[gazemodel_sample$age_group == 18]` adults from our Lifespan study (mean age = `r gazemodel_sample$age_mean[gazemodel_sample$age_group == 18]` years, SD = `r gazemodel_sample$age_sd[gazemodel_sample$age_group == 18]`, range = `r gazemodel_sample$age_min[gazemodel_sample$age_group == 18]` - `r gazemodel_sample$age_max[gazemodel_sample$age_group == 18]`, `r gazemodel_sample$n_females[gazemodel_sample$age_group == 18]` female).
Adults were recruited over *Prolific* [@palan2018prolific].
Since developmental change was minimal in our adult sample (see Lifespan study) and the cognitive models were computationally heavy, we decided to only include the first 50 adults that had completed the study.

@@

\begin{equation}
    P(\mathbf{\theta} | x_c, \alpha_l, \alpha_r) \propto P(x_c | \alpha_l, \alpha_r, \mathbf{\theta})P(\mathbf{\theta})
\end{equation}

where $\mathbf{\theta}$ is an individual's cognitive ability to locate the focus of the agent's attention, $x_c$ is the coordinate the participant clicked, and $\alpha_l$ and $\alpha_r$ are the pupil angles for the left and right eye, respectively. The pupil angle is defined as the angle between a line connecting the center of the eye to the pupil and a line extended vertically downward from the pupil.[^3]

[^3]: This model mirrors the logic of the TANGO programming code.
    In the online experiment, we read out the center point coordinates of the target and the agent's eyeball (i.e., the SVG coordinates).
    We then calculate a line between these two points: this is our gaze vector.
    Now, knowing the eyeball radius, we calculate the point of intersection at which the gaze vector meets the eyeball boundary.
    Finally, the agent's pupil moves from the center of the eyeball along the gaze vector to the intersection point.
    This way, the agent is animated to "look at" the target.
    In the gaze model, we assume participants go through these steps in reverse order.

>>>>

abstract: |
  Observing eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze following in infancy has been thoroughly studied, we know little about the developmental trajectory throughout childhood and adolescence. The present study examines gaze understanding across the lifespan. We formalize the process of gaze understanding in a computational cognitive model that allows us to conceptualize individual differences in a psychologically meaningful way. The key parameter in our model is an inferential component, which describes how accurately a participant infers a target location based on the agent's gaze. We test a fundamental assumption of our gaze model by studying vector estimation in a non-social setting. Lastly, we explore the relationship between other social-cognitive abilities and gaze understanding. We found a steep learning curve in gaze understanding during the preschool years. In these years, children become more precise in locating the attentional focus of an agent. Precision levels then stay comparably stable, with a minor decay toward older adulthood. Our gaze model estimates reliable individual parameter values and recovers signature patterns in the data. Furthermore, we found that gaze understanding is associated with non-social vector estimation and perspective-taking but not with other Theory of Mind tasks. This work illustrates how the combination of reliable tasks and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan"
wordcount         : "xxx max 8000"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes   :
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
---

@@

**Acknowledgements:** We thank Jana Jurkat for her help with data collection and participant recruitment. We would also like to thank Steven Kalinke for his technical programming support. We thank all the children, caregivers, and adults who participated in the study.

**Funding:** This study was funded by the Max Planck Society for the Advancement of Science, a noncommercial, publicly financed scientific organization (no grant number). Manuel Bohn was supported by a Jacobs Foundation Research Fellowship.

**Ethical statement:** The studies were approved by an internal ethics committee at the Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany. Informed consent was obtained from all individual participants or their legal guardians.

@@

# Research highlights (to be placed before the abstract)

-	Gaze understanding develops beyond infancy. Young adults reach the highest precision levels in localizing an attentional focus, while older adults perform slightly less precisely. 
-	We present a computational model as a mechanistic description of gaze understanding. The model explains individual differences and recovers signature patterns in the data.
-	Our model assumes that gaze understanding relies on vector estimation. To test this assumption experimentally, we designed a non-social vector estimation task. 
-	We found an association between gaze understanding and vector estimation, as well as Level 2 perspective-taking. Other Theory of Mind tasks did not correlate.

# Introduction
How do humans learn about their environment and navigate through their social surroundings? A promising approach to extract information from the environment is through following the gaze of others. Recognizing the attentional focus of other agents also helps in building common ground and is vital for communicative interactions and shared activities [@tomasello2007reliance].

Existing studies operationalize gaze following as the ability to follow another agent’s line of sight. As one of the most fundamental social-cognitive abilities, it has been extensively studied in infancy and early childhood. The traditional, often-used paradigm to study gaze following looks as follows [e.g., @gredeback2010development]: The experimenter sits in front of the infant, with one object to their left and one to the right. The experimenter looks directly at the infant before shifting her head and eyes to one of the two objects. Infants’ looking times or dichotomous object choices are then recorded. Analyses traditionally focus on the average age at which children reach an above-chance performance (i.e., looking at the gaze-cued target object). According to this research, infants as young as six months can attune their gaze to that of another agent  [@dentremont1997demonstration]. At the end of their first year of life, infants can follow gaze to locations outside their current visual field and move themselves to gain proper perceptual access [@moll200412; @deak2000effects].

At first glance, the debate about infants’ gaze-following abilities seems to be settled. On further view, however, some methodological choices in the experimental paradigms as well as the scope of infants’ abilities seem to offer room for further exploration. We see the following challenges: (1) Studies focus on infancy, leaving aside further potential developments in childhood and adolescence; (2) Gaze cues are often accompanied by a head movement in the same direction, which makes it hard to isolate the signaling power of the eyes; (3) Often, infants’ looking times are taken as a benchmark for passing a task, while it is uncertain how this would translate into active behavior; (4) Presenting two objects drastically limits the complexity of the visual input, especially when compared to our daily, visually cluttered environment; (5) Behavior is grouped into passing/failing or a comparison against chance, ignoring the accuracy of children; (6) Analyses focus on group-level averages, overlooking variation in the sample.

Adding to the first point, while the emergence of gaze following has been well established, less is known about the developmental trajectory throughout childhood and adolescence. One possibility is that our social-cognitive ability in question is fully developed once emerged in infancy. However, many cognitive abilities develop with age [e.g., working memory, @gathercole2004structure]. Similarly, visual processing appears to improve with age. Therefore, children could potentially improve in gaze following, fine-tuning the performance of the already existing skill. We believe that the development of gaze understanding is, most likely, not complete merely because infants can differentiate between two objects, one which has been gazed at. In daily life, we often face more nuanced, fine-grained gaze cues and must differentiate between a virtually unlimited number of potential targets.

Let us now turn to the point that often the experimenter shifts their eyes and head in synchrony [e.g., @behne2005oneyearolds]. Instead of pointing towards gaze understanding, a critic could claim that the results can be explained by face direction alone. A handful of studies approached this potential confound by separately manipulating head and eye movement. @brooks2002importance implemented a comparison between eye and head orientation and found that 14-month-olds were sensitive to open versus closed eyes. @moore1997role showed that 9-month-olds followed an agent’s gaze more when it was accompanied by a dynamic head turn in comparison to a static head turn. Investigating the 'cooperative eye hypothesis', @tomasello2007reliance implemented six conditions, in which an experimenter oriented towards the ceiling with their eyes only, head only (eyes closed), both head and eyes, or neither. They found that human infants relied more on the eye movement, while chimpanzees paid more attention to the head movement. Results of @corkum1995development, @lempers1979young, and @lempers1977development suggest that infants, at least until 19 months, struggle when eye and head direction diverge. This underlines how important the movement of the head is when interpreting the gaze of others and that the scope of infants' gaze understanding might be less comprehensive than previously assumed.

Next, let us consider the classical response measure of collecting looking times. Though previous studies using the above-mentioned methods suggest that young infants can align their visual attention to another’s line of sight, it does not necessarily include understanding the intentions or the content of the other agent’s gaze. Infants could simply attune their orientation or be attracted by others’ gaze without processing what exactly the other is seeing [cf. Butterworth & Jarrett's ecological and geometric mechanism, @butterworth1991minds]. Arguably, this is not sufficient to claim an understanding of visual perspectives, because the “target” or “object of representation” is not necessarily specified [@perner2003perspective, p. 358]. To fulfill this criterion, @moll2006level used an active behavioral choice as a response measure and tested infants' understanding of the intentions behind gaze. Infants interactively searched for the target object, which was either occluded or visually accessible. They found that 24-month-olds adaptively handed over the target object, while 18-month-olds still struggled to complete this task. In a hiding game with two search locations, @povinelli1997exploitation found that two-year-olds performed at chance level when using gaze as a cue to locate the reward, while three-year-olds passed the task. However, in a similar object choice paradigm with two containers, @behne2005oneyearolds investigated whether infants understand the communicative intent behind pointing and gaze cues. In contrast to @povinelli1997exploitation, they found that already 14-month-olds used the agent’s cues to select an object. In conditions with absent-minded ‘cues’, infants performed around chance. The specific scope of children’s gaze understanding and whether correct looking translates into meaningful active behavior deserve further investigation.

Finally, presenting two objects, focusing on group-level analysis, and passing versus failing a task drastically limit the level of detail with which we can gauge children’s gaze understanding. Undoubtedly, children can vary in the onset of their social-cognitive abilities: these skills develop over time, with experience, and on a continuum. @birch2017perspectives argued that individuals can differ in their frequency (or propensity) and accuracy of applying social cognition in their daily lives. Measuring individual differences in this domain is especially important to understand the underlying processes and to disentangle the impact of environmental influences (scaffolding) and other cognitive abilities [@birch2017perspectives].

@@

With our study, we aim to deepen our knowledge about the development and mechanisms behind gaze understanding. We study the differentiation of this social-cognitive ability by using a study design with subtle gaze cues without head rotation. While we expect the younger children to be able to follow gaze, we aim at assessing the differentiation of their social-cognitive ability. Our goal was *not* to establish the youngest age at which children understand gaze cues. Rather, we wanted to examine *how* that ability changes with age. In our study, we focused on the communicative intents of gaze: we asked children to locate a target by following an agent’s gaze. While language demands are low, participants have to actively respond and, therefore, make use of the presented gaze cue. By measuring a continuous, active behavior, we can capture meaningful variation between individuals. We strive to embrace and explain diversity in gaze understanding and use individual differences to inform what happens on a process level.

The present paper is structured in the following way. In Study 1, we explore gaze understanding across the lifespan. We assess the developmental trajectory and individual differences in this fundamental social-cognitive ability in a sample of 3- to 80-year-olds. In Study 2, we propose a computational cognitive model of gaze understanding. This model helps us to explain where individuals vary and which mechanisms participants use to locate an agent’s attentional focus. In Study 3, we revisit a fundamental assumption of our computational gaze model by experimentally isolating the vector estimation component of gaze understanding. Furthermore, we investigate the relationship between gaze understanding and other (social-) cognitive processes.

@@

@@

We collected data from a remote child, teenager and adult sample, reaching from 3 to 80 years of age (see Supplements for further details). The child and teenager sample consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group < 18])` participants. We recruited participants via an internal database consisting of families living in Leipzig, Germany, who volunteered to participate in child development studies and indicated an interest in online studies. Participants came from ethnically homogeneous, mixed socioeconomic backgrounds with mid to high parental education levels. They lived in an industrialized, urban Central-European context in a mid-size German city (∼600,000 inhabitants; median individual monthly net income ~ 1,600€ as of 2021). Most were raised monolingually in a nuclear two-generational family setting. Information on demographics and socioeconomic status was not formally recorded on a participant level.

Adults were recruited via *Prolific* [@palan2018prolific]. *Prolific* is an online participant recruitment service from the University of Oxford with a predominantly European and US-American subject pool. Participants consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group >= 20])` English-speaking adults who reported to have normal or corrected-to-normal vision. For completing the study, subjects were paid above the fixed minimum wage (on average £10.00 per hour).

@@

We used the continuous version of the TANGO [@prein2023tango]. The task was presented as an interactive web application (live demo [https://ccp-odc.eva.mpg.de/tango-demo/](https://ccp-odc.eva.mpg.de/tango-demo/.); source code <https://github.com/ccp-eva/tango-demo>). The TANGO showed satisfactory internal consistency and retest reliability (with reliability estimates *Pearson's r* ranging from .7 to .8 for the continuous task version; [@prein2023tango]).

@@

To estimate the effect of age on gaze understanding, we fit GLMMs that make different assumptions about the developmental trajectory, modeling the relationship as linear, quadratic, or cubic. In addition, we applied a Gaussian Process model [@burkner2017brms] and examined nonlinearity. The models predicted imprecision by age (continuous), aggregated across trials and modeled as a `lognormal` distribution.[^1] The unit of imprecision was counted in target widths, i.e., a participant with imprecision of 1 clicked on average one target width to the left or right of the true target center. We inspected the posterior distributions (mean and 95% Credible Interval (CrI)) for the age estimates and compared models using leave-one-out cross-validation (LOO), the widely applicable information criterion (WAIC), and model weights [@vehtari2017practicala].

[^1]: Originally, our models were fitted on a trial-by-trial basis and implemented the following structure: `performance ~ age + symmetricPosition + trialNr + (1 + symmetricPosition + trialNr | subjID)`. However, the Gaussian Process model was computationally heavy. To circumvent issues of convergence, we simplified the model structure, aggregated data on a subject level, and included only age as an effect. See Supplements for a comparison between the original and the here-reported model structures. The model predictions did not differ notably.

@@

(ref:figlab1) **Developmental trajectory of gaze understanding across the lifespan**. Performance is measured as imprecision, i.e., the absolute distance between the target's center and the participant's click (averaged across trials). The unit of imprecision is counted in the width of the target, i.e., a participant with imprecision of 1 clicked on average one target width to the left or right of the true target center. Grey dots show the mean performance for each subject averaged across trials. Blue lines show 100 draws from the expectation of the posterior predictive distribution (i.e., conditional expectation) of the Gaussian Process model, with its mean predicted developmental trajectory as a solid black line. Vertical, black, dashed lines show the locations of the most prominent changes according to our Bayesian change point analysis.

```{r fig1, include = T, out.width = "100%", fig.align = "center", fig.cap = "(ref:figlab1)"}
knitr::include_graphics(c("../figures/lifespan_plot.png"))
```

@@

High levels of variation pointed to individual differences in all age groups (overall imprecision mean = `r lifespan_variation$imprecision_mean`, sd = `r lifespan_variation$imprecision_sd`, range = [`r lifespan_variation$imprecision_min` - `r lifespan_variation$imprecision_max`]). For example, there were some children who were more accurate than the average adult.

In our model comparison, we found clear evidence for a non-linear development in gaze understanding across the lifespan. Compared against linear, quadratic, and cubic models, the Gaussian Process model showed the best model fit, demonstrated by the greatest model weight and smallest WAIC values. For the imprecision in gaze understanding, the standard deviation (*b* = `r lifespan_gp$estimate[lifespan_gp$param == "sdgp(gpage_centered)"]`, 95% CrI [`r lifespan_gp$l_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`; `r lifespan_gp$u_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`]) and visual inspection indicated nonlinearity (see Figure \@ref(fig:fig1)).

Going one step further, we investigated the most prominent change points in the data. The Bayesian change point analysis revealed `r lifespan_beast$trend$ncp_mode` (with `r lifespan_beast$trend$ncpPr[lifespan_beast$trend$ncp_mode + 1] * 100 %>% round(2)`% probability) major shifts in gaze understanding during the lifespan. The change points occurred at the following ages: `r lifespan_cps$cp[1]` years (95% CrI [`r lifespan_cps$lowerCrI[1]`; `r lifespan_cps$upperCrI[1]`]); `r lifespan_cps$cp[2]` years (95% CrI [`r lifespan_cps$lowerCrI[2]`; `r lifespan_cps$upperCrI[2]`]); `r lifespan_cps$cp[3]` years (95% CrI [`r lifespan_cps$lowerCrI[3]`; `r lifespan_cps$upperCrI[3]`]); `r lifespan_cps$cp[5]` years (95% CrI [`r lifespan_cps$lowerCrI[5]`; `r lifespan_cps$cp[4]` years (95% CrI [`r lifespan_cps$lowerCrI[4]`; `r lifespan_cps$upperCrI[4]`]); `r lifespan_cps$upperCrI[5]`]); and finally, at `r lifespan_cps$cp[6]` years (95% CrI [`r lifespan_cps$lowerCrI[6]`; `r lifespan_cps$upperCrI[6]`]). In other words: we found a very rapid initial improvement in early childhood, followed by a long period of minor, very slow change with slightly increasing levels of imprecision toward the eldest in our sample.

@@

We investigated the shape of change in gaze understanding across the lifespan. By applying the exact same task for the entire age range, we could directly compare gaze understanding in all ages. Therefore, we circumvented methodological challenges that often arise when trying to compare data across ages from qualitatively and quantitatively different study tasks.

@@

Previous studies found that children start to follow gaze in the second half of their first year of life [@moll200412]. In our sample, three-year-olds were still rather imprecise in their gaze understanding ability. How can we explain this divergence? First of all, we used subtle eye movements as cues. Many existing studies let the agents move eye and head in parallel [@behne2005oneyearolds; @povinelli1997exploitation], therefore establishing a confound with more salient head movement. Relying exclusively on eye movements might be more difficult for children than presenting them with a combined eye and head orientation [@carpenter1998social]. Furthermore, our study required participants to (1) precisely follow an agent's gaze, (2) interpret this as a cue, and then (3) make use of this cue to guide their own behavior (i.e., touching the screen at the cued location). We measured an active location choice instead of an attention orientation. It is conceivable that three-year-olds followed the agent's gaze but were still learning to translate this understanding into precise, active behavior.

Regarding the sample of elderly adults, we expect a sampling bias to be present [@bethlehem2010selection; @gosling2004should; @remillard2014systematic]. First, certainly not all older people have a working high-speed Internet connection or are knowledgeable and trained in its use. Second, it takes a certain amount of independence and motivation to participate in *Prolific* studies. The elderly who know how to participate in *Prolific* studies might show greater cognitive fitness and flexibility compared to their offline counterparts. Therefore, a representative sample might show a greater age decline in gaze understanding compared to our reported sample. In addition, it is noteworthy that older people might be more likely to suffer from visual impairments. Even though we filtered participants to only include normal- to correct-to-normal vision, we cannot guarantee that our participants showed no symptoms of reduced vision.

@@

Our lifespan study showed that gaze understanding develops throughout childhood, and variation between individuals appears in all age groups. The TANGO has previously been shown to reliably capture inter-individual differences in gaze understanding [@prein2023tango]. The variation between participants is thus likely genuine and not due to random noise. Now, we aim to understand what explains the developmental change and the variation across participants on a process level. We present a theory of gaze understanding that explains how children process the available information (i.e., the agent's eyes) to make inferences about the agent's gaze and its attentional focus, which leads them to identify the target position. We formalize this inference process in a computational cognitive model.

@@

We seek to explain how participants solved the TANGO. We designed a computational cognitive model that replicates a schematic representation of how participants make inferences in the task's context (i.e., model of the task and not the data). The fundamental assumption of this gaze model is that participants use all available gaze information to infer the target location.

@@

In addition, we included `r gazemodel_sample$n_total[gazemodel_sample$age_group == 18]` adults from our Lifespan study (mean age = `r gazemodel_sample$age_mean[gazemodel_sample$age_group == 18]` years, SD = `r gazemodel_sample$age_sd[gazemodel_sample$age_group == 18]`, range = `r gazemodel_sample$age_min[gazemodel_sample$age_group == 18]` - `r gazemodel_sample$age_max[gazemodel_sample$age_group == 18]`, `r gazemodel_sample$n_females[gazemodel_sample$age_group == 18]` female). Adults were recruited over *Prolific* [@palan2018prolific]. Since developmental change was minimal in our adult sample (see Lifespan study) and the cognitive models were computationally heavy, we decided to only include the first 50 adults who had completed the study.

@@

\begin{equation}
    P(\theta | x_c, \alpha_l, \alpha_r) \propto P(x_c | \alpha_l, \alpha_r, \theta)P(\theta)
\end{equation}

where $\theta$ is an individual's cognitive ability to locate the focus of the agent's attention, $x_c$ is the coordinate the participant clicked, and $\alpha_l$ and $\alpha_r$ are the pupil angles for the left and right eye, respectively. The pupil angle $\alpha$ is defined as the angle between a line connecting the center of the eye to the pupil and a line extended vertically downward from the center of the eye (see \@ref(fig:fig2)A).[^3]

[^3]: This model mirrors the logic of the TANGO programming code. In the online experiment, we read out the center point coordinates of the target and the agent's eyeball (i.e., the SVG coordinates). We then calculate a line between these two points: this is our gaze vector (acting in the functionally same way as a pupil angle). Now, knowing the eyeball radius, we calculate the point of intersection at which the gaze vector meets the eyeball boundary. Finally, the agent's pupil moves from the center of the eyeball along the gaze vector to the intersection point. This way, the agent is animated to "look at" the target. In the gaze model, we assume participants go through these steps in reverse order.
